---
title: "Exploratory Data Analysis - US Crime"
author: "Hannah Pavlovich"
date: "2024-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r practice}
### GOOD PRACTICES ###
library(outliers)
library(ggplot2)
library(gridExtra)
library(randomcoloR)
library(car)
library(DAAG)
rm(list = ls())
set.seed(42) # set a seed so it everything be reproducible
```


Criminologists are interested in the effect of punishment regimes on crime rates. To this effect, data has been aggregated on 47 states for various demographic indicators. The Response variable is crime rate: number of offenses per 100,000 population in 1960.

The following performs Exploratory Data Analysis on the set provided, including outlier detection and removal.

#### Check for normality in the model with qq-norm.

```{r}
### Read in the data
crimedata <- read.table("data/uscrime.txt", stringsAsFactors = FALSE, header = TRUE)
## Run crime through qqnorm
qqnorm(crimedata$Crime, col = "darkblue")
qqline(crimedata$Crime, col = "red")
## create a histogram
hist(crimedata$Crime, xlab = "Crime", main = "Histogram of Crime")
```

Looking at our data on the crime, it is not normal. The histogram is
skewed to the left, and the QQ Plot shows deviances from the normal
line.

#### Histograms, scatterplots, and correlation matrix for each variable.

```{r message=FALSE, warning=FALSE}

## Function to create Histograms
hist_plot <- function(i){
  ggplot(crimedata, aes(x = crimedata[,i])) + 
    geom_histogram( color="black", fill= randomColor()) +  xlab(names(crimedata[i]))
}

## Return Histograms

  do.call(grid.arrange, lapply(1:15, hist_plot))
  

```

According to the histograms, most of the data is skewed to the left with
a fairly normal distribution.

Ed, the mean years of schooling, does not follow this trend, and seems to
have an almost bi-modal distribution.

Pop, the state population, is completely left skewed with no normal
distribution.

Wealth is the only histogram that is heavier on the right side.

I next examined the scatterplots to see what the relationships are
between the variables. We are using multiple variables to predict the
crime rate, so it is important to see how the predictors relate to one
another

```{r message=FALSE, warning=FALSE}

## Function to create scatter plots
scat_plot <- function(i){
  ggplot(crimedata, aes(x = crimedata[,i], y = Crime)) + 
    geom_point(weight = 1.5, color = randomColor()) +  xlab(names(crimedata[i]))
}

## Return scatterplots

  do.call(grid.arrange, lapply(1:15, scat_plot))

```

We can see some linear correlation especially with Po (per capita
expenditure on police protection) and wealth. There also seems to be a
non-linear relationship between crime and Pop (state population in 1960
in hundred thousands), NW (percentage of nonwhites in the population),
and U1 and U2 (unemployment rates of urban males)

#### Outlier Detection

To check for outliers, the first step is to do an EDA of the data. I
begin with the summary.

```{r}
summary(crimedata$Crime)
```

First inspection shows a Max that is 1,100 points higher than the Median
and 900 points higher than the third quantile. This suggests there are
upper outliers. The lower numbers are less extreme.

Next, I make a boxplot to inspect the data

```{r pressure, echo=FALSE}
ggplot(data = crimedata, aes(y = Crime)) + geom_boxplot()
```

The boxplot is similar to our summary. By visual inspection, I can posit
that there are at least two upper outliers.

My last EDA is with a histogram

```{r}
ggplot(data = crimedata, aes(x=Crime)) + geom_histogram(binwidth = 150, 
      color = "black", fill = "lightblue") + 
      geom_vline(xintercept = mean(crimedata$Crime), 
      color = "red", linetype = "dashed", size = 1
)
```

The histogram above plots the Crime data against count, with the red
dashed line showing the mean of Crimes. Again, we can see that this
histogram is right-side heavy, suggesting outliers.

Next, I will apply the grubbs test to see if these points are actually
outliers, and to see if they will change the data.

```{r}
print("Grubbs Test for Highest Value:")
grubbs.test(crimedata$Crime)

print("Grubbs Test for Lowest Value")
grubbs.test(crimedata$Crime, opposite = TRUE)

```

Above, I ran the Grubbs Tests to check for outliers on the highest and
lowest ends. According to this test, our high point 1993 is NOT an
outlier with $\alpha$ = 0.05.

By quick inspection, plots against Po1 and Po2 imply some sort of
linearity. While not getting too deep into this, I can reject the
hypothesis that an outlier exists.

#### Correlation

```{r}
res <- cor(crimedata[,1:15], method="pearson")
corrplot::corrplot(res, method= "square", order = "AOE", tl.pos = 'n')
```

From the above, we have a better understanding of how the predictor
variables are correlated. Some of them have very strong correlations
while many of them are near-zero.